---
title: "Data Management for Clinical Researchers"
author: "Albert Cobos"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook: default
---


```{r setup, include=FALSE, eval = TRUE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = NA)  # class.output = "custom-output"

```

# Preface {-}

This is an introductory book on data management for clinical researchers. The ability to manage data is a fundamental skill for data analysts, because only rarely the initial state of a dataset is adequate to perform the desired analysis. Rather, preparation tasks are usually needed to transform the data so that analysis functions can be used. In fact, these tasks tend to be more complex, more prone to errors, and much more time consuming, than the analysis itself. 

All data management tasks can be accomplished using base R. However, several R packages make them easier. Among these, packages `dplyr`, `tidyr`, `stingr` and `forcats` will be used in this book. These (and other) packages form a collection known as the [tidyverse](https://www.tidyverse.org/), which is integrated in a package of the same name (`tidyverse`). Once installed, you can load it with:

```{r}
library(tidyverse)
```

\

And that's all you need to start!

```{r 0-clean, echo=FALSE, eval=TRUE}
rm(list = ls())
```
